---
author: "Centre for Population Genomics"
date: "`r as.character(paste(Sys.time(), Sys.timezone()))`"
output:
  html_document:
    theme: cosmo
    toc: true
    code_download: true
    code_folding: show
  rmdformats::material:
    highlight: kate
params:
  title: ""
  bucket_suffix: "test"
  gender_tsv: "gender.tsv"
  age_csv: "age.csv"
  qc_csv: "qc.csv"
title: "Exploration of QC metrics for TOB-WGS"
---

```{r knitr_opts, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Introduction

Here we're exploring QC metrics, metadata, and contents of Google Cloud Storage
buckets associated with the TOB-WGS project.

```{r load_pkgs}
library(tidyverse)
library(fs)
library(glue)
library(here)
library(assertthat)
library(reactable)
library(ggrepel)
library(sessioninfo)
library(DT)

# for GCS authentication
library(googleCloudStorageR)
library(gargle)
```

```{r funcs}
guess_file_type <- function(x) {
  dplyr::case_when(
    grepl("\\.bam$", x, ignore.case = TRUE) ~ "BAM",
    grepl("\\.bai$", x, ignore.case = TRUE) ~ "BAMindex",
    grepl("\\.cram$", x, ignore.case = TRUE) ~ "CRAM",
    grepl("\\.crai$", x, ignore.case = TRUE) ~ "CRAMindex",
    grepl("\\.fastq.gz$", x, ignore.case = TRUE) ~ "FASTQ",
    grepl("\\.fastq$", x, ignore.case = TRUE) ~ "FASTQ",
    grepl("\\.fq$", x, ignore.case = TRUE) ~ "FASTQ",
    grepl("\\.fq\\.gz$", x, ignore.case = TRUE) ~ "FASTQ",
    grepl("manifest\\.txt$", x, ignore.case = TRUE) ~ "Manifest",
    grepl("\\.md5$", x, ignore.case = TRUE) ~ "MD5",
    grepl("md5\\.txt$", x, ignore.case = TRUE) ~ "MD5txt",
    grepl("\\.vcf$", x, ignore.case = TRUE) ~ "VCF_unz",
    grepl("\\.g\\.vcf\\.gz$", x, ignore.case = TRUE) ~ "GVCF",
    grepl("\\.vcf\\.gz$", x, ignore.case = TRUE) ~ "VCF",
    grepl("\\.tbi$", x, ignore.case = TRUE) ~ "VCFindex",
    grepl("\\.csv$", x, ignore.case = TRUE) ~ "CSV",
    grepl("\\.json$", x, ignore.case = TRUE) ~ "JSON",
    TRUE ~ "OTHER")
}

gcs_list_objects2 <- function(b) {
  googleCloudStorageR::gcs_list_objects(bucket = b,
                                        detail = "summary") %>%
    dplyr::as_tibble() %>%
    dplyr::mutate(name = glue::glue("gs://{b}/{name}"),
                  size = sub(" bytes", "", size), # else returns NA
                  size = fs::as_fs_bytes(size),
                  ftype = guess_file_type(name))
}

# The given observation x is considered an outlier
# (see https://statsandr.com/blog/outliers-detection-in-r/).
# Using the m parameter for cases where there are simply too many outliers; the
# larger the value of m, the fewer values are considered outliers.
is_outlier <- function(x, m = 1.5) {
  assert_that(is.numeric(x), is.numeric(m))
  (x < quantile(x, 0.25) - m * IQR(x)) |
    (x > quantile(x, 0.75) + m * IQR(x))
}
```


```{r vars}
bucket <- glue("cpg-tob-wgs-", params$bucket_suffix)
date <- "2021-05-21"
```

## WGS GVCF files

- Let's first explore the contents of the
  `r glue("{bucket}")` bucket:

```{r list_contents1}
scope <- c("https://www.googleapis.com/auth/cloud-platform")
token <- gargle::token_fetch(scopes = scope)
googleCloudStorageR::gcs_auth(token = token)
obj_list <- gcs_list_objects2(b = bucket)
```

```{r list_contents2}
count(obj_list, ftype) %>% knitr::kable(caption = glue("Count of file types in {bucket}."))
gvcf <-
  obj_list %>%
  dplyr::filter(ftype == "GVCF") %>%
  dplyr::mutate(basename = basename(name),
                tobid = sub(".g.vcf.gz", "", basename),
                batch = sub("batch(.)", "\\1", basename(dirname(name)))) %>%
  dplyr::select(tobid, fullname = name, batch, size, basename)
```

- Total of **`r nrow(obj_list)`** files, consisting of **`r nrow(gvcf)`**
  GVCF files:

```{r gvcf_list}
gvcf_tab <- gvcf %>%
  dplyr::mutate(size = as.character(size),
                n = dplyr::row_number()) %>%
  dplyr::select(n, fullname, batch, size) %>%
  reactable::reactable(
    pagination = FALSE, highlight = TRUE, height = 500, searchable = TRUE,
    filterable = TRUE, bordered = TRUE, fullWidth = FALSE,
    columns = list(
      n = colDef(minWidth = 60),
      fullname = colDef(minWidth = 420),
      batch = colDef(minWidth = 60),
      size = colDef(minWidth = 80)
    )
  )

htmlwidgets::prependContent(
  gvcf_tab,
  htmltools::h2(class = "title", glue("GVCF files in {bucket} (Total: {nrow(gvcf)})")))
```

```{r gvcf_file_sizes}
theme_set(theme_bw())

gvcf %>%
  summarise(
    min = fs::as_fs_bytes(min(size)),
    max = fs::as_fs_bytes(max(size)),
    q1 = fs::as_fs_bytes(quantile(size, 0.25)),
    median = fs::as_fs_bytes(median(size)),
    q3 = fs::as_fs_bytes(quantile(size, 0.75)),
    total = fs::as_fs_bytes(sum(size)),
    .groups = "drop") %>%
  tidyr::pivot_longer(cols = dplyr::everything()) %>%
  knitr::kable(caption = glue("Size metrics for {nrow(gvcf)} GVCF files in {bucket}."))

# jitter object for proper labels
jitter_pos1 <- ggplot2::position_jitter(width = 0.10, height = 0, seed = 321)

gvcf %>%
  dplyr::mutate(
    outlier = dplyr::if_else(is_outlier(size, 3),
                             glue::glue("{tobid} ({size})"), NA_character_)) %>%
  ggplot(aes(x = batch, y = size)) +
  geom_violin(fill="lightblue") +
  geom_point(position = jitter_pos1, shape = 21, color = "darkblue", fill = "red") +
  ggrepel::geom_text_repel(position = jitter_pos1, aes(label = outlier),
                           na.rm = TRUE, hjust = -0.3) +
  scale_y_continuous(labels = scales::comma, breaks = scales::breaks_pretty(10)) +
  theme(panel.grid.minor = element_blank()) +
  ggtitle(glue::glue("Size (bytes) for {nrow(gvcf)} GVCF files in {bucket}."))
```

## QC metrics

We have access to CSV files with summaries of sample QC metrics. Additional
files have been provided containing:

- the standard deviation of CRAM insert size, as calculated by Picard's
  CollectInsertSizeMetrics;
- the gender of each sample;
- the age of each sample;

```{r qc_plot_prep}
qc1 <- params$qc_csv %>% 
  readr::read_csv()

qc_gender <- params$gender_tsv %>% 
  read_tsv(
    skip = 2,
    col_names = c("sample_id", "sample.sample_name", "gender",
                  paste0("nanodrop_", c("conc", "260/280", "230/280")),
                  paste0("qubit_", c("assay_conc", "stock_conc")),
                  "volume", "location")) %>%
  dplyr::filter(!is.na(sample_id)) %>%
  dplyr::select(sample = sample.sample_name, gender)

qc_age <- params$age_csv %>%
  readr::read_csv(col_types = cols(.default = "c", age = "d")) %>%
  dplyr::select(sample = TOBIID, age)

p_data <- qc1 %>%
  dplyr::select(sample = sample.sample_name,
                `freemix contamination` = raw_data.FREEMIX,
                `chimeras pct` = raw_data.PCT_CHIMERAS,
                `duplicate pct` = raw_data.PERCENT_DUPLICATION,
                `insert size median` = raw_data.MEDIAN_INSERT_SIZE,
                `coverage median` = raw_data.MEDIAN_COVERAGE) %>%
  tidyr::pivot_longer(cols = -sample, names_to = "metric") %>%
  dplyr::left_join(qc_gender, by = "sample") %>%
  dplyr::left_join(qc_age, by = "sample") %>%
  dplyr::left_join(gvcf %>% dplyr::select(sample = tobid, batch), by = "sample") %>%
  dplyr::mutate(batch = glue("batch {batch}"))
```

```{r plot_all_metrics, fig.height=18, fig.width=12}
# jitter object for proper labels
jitter_pos2 <- ggplot2::position_jitter(width = 0.10, height = 0, seed = 321)

p_data %>%
  dplyr::group_by(metric) %>%
  dplyr::mutate(
    value_rounded = round(value, 3),
    outlier = if_else(is_outlier(value),
                      glue("{sample} ({value_rounded})"),
                      NA_character_)) %>%
  ggplot(aes(x = "", y = value)) +
  geom_point(position = jitter_pos2, shape = 21, colour = "blue") +
  ggrepel::geom_text_repel(position = jitter_pos2, aes(label = outlier),
                           na.rm = TRUE, hjust = 9.3) +
  geom_violin(fill = NA) +
  facet_grid(metric ~ batch, scales = "free") +
  scale_y_continuous(breaks = scales::breaks_pretty(8)) +
  ggtitle(glue("QC metrics by batch for ",
               "{length(unique(p_data$sample))} TOB-WGS samples.")) +
  theme(strip.text.x = element_text(face = "bold.italic"),
        strip.text.y = element_text(size = 12,face = "bold.italic"))
```

## Session Info

```{r session_info, echo=FALSE}
si <- sessioninfo::session_info(include_base = TRUE)
si_pl <- unclass(si$platform) %>% as_tibble() %>% t()
si_pkg <- unclass(si$packages) %>%
  dplyr::as_tibble() %>%
  dplyr::select(package, version = ondiskversion, datestamp = date)
dplyr::tibble(var = rownames(si_pl),
              value = si_pl[, , drop = TRUE]) %>%
  knitr::kable(caption = "Platform information.")
si_pkg %>%
  DT::datatable(filter = list(position = "top", clear = FALSE, plain = TRUE),
                rownames = FALSE, extensions = c("Scroller"),
                options = list(scroller = TRUE, scrollX = TRUE, scrollY = 200,
                               dom = 'Bfrtip'))
```
